From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting

### 2024-03-15 paper/paper.qmd
Update paper

---
title: "Preparing your manuscript"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Florian Börgel
    affiliations: 
      - ref: IOW
    orcid: 0000-0003-2402-304X
    email: florian.boergel@io-warnemuende.de
    corresponding: true

@@

abstract: |
  The abstract (1) states the nature of the investigation and (2) summarizes the important conclusions. The abstract should be suitable for indexing. Your abstract should:

plain-language-summary: |
  A Plain Language Summary (PLS) can be an incredibly effective science communication tool. By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience. A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.  A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline. Read our tips for creating an effective PLS.
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

@@

LSTM networks, an evolution of the classical Recurrent Neural Networks (RNNs), have shown stability and efficacy in sequence-to-sequence predictions, such as using climatic indices for rainfall estimation or long-term hydrological forecasting. However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel. Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff.

TODO: More about ConvLSTM

In the following we will use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions. The Baltic Sea serves as a challenging example due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).

As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers. More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface. The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input (Meier and Doescher, 2002). Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation. This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea. In the following we show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM represent a robust way to predict river runoff for any given period of time using only atmospheric forcing.

In this work, we present a ConvLSTM architecture as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.

@@

>>>>

---
title: "From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Florian Börgel
    affiliations: 
      - ref: IOW
    orcid: 0000-0003-2402-304X
    email: florian.boergel@io-warnemuende.de
    corresponding: true

@@

abstract: |
    a
plain-language-summary: |
    b
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Climate Dynamics
keep-tex: true
date: last-modified
---

@@

LSTM networks, an evolution of the classical Recurrent Neural Networks (RNNs), have shown stability and efficacy in sequence-to-sequence predictions, such as using climatic indices for rainfall estimation or long-term hydrological forecasting. However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel. Reconizing this limitation [@shi2015] proposed a convolutional LSTM (ConvLSTM) architectures, which combines the strengths of both LSTM and CNN. The ConvLSTM network has been proven useful for precipitation nowcasting [@shi2015] or for river runoff forecasting [@ha2021; \@niu2020].

In the following we will show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM also represent a robust way to predict multiple rivers at once for any given period using only atmospheric forcing. In this work we use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions. The Baltic Sea serves as a challenging example due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure). As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers. More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface. The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input (Meier and Doescher, 2002). Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation. This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea.

In this work we will, we present a ConvLSTM architecture that is able to predict daily river runoff for 97 rivers across the Baltic Sea catchment.

@@

### Runoff data used for training

### Atmospheric Forcing

### 2024-03-15 paper/paper.tex
Update paper

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Preparing your manuscript},
  pdfauthor={Florian Börgel; Sven Karsten},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Geophysical Research Letters}

@@

\begin{document}
\title{Preparing your manuscript}

@@

\begin{abstract}
The abstract (1) states the nature of the investigation and (2)
summarizes the important conclusions. The abstract should be suitable
for indexing. Your abstract should:
\end{abstract}

\section*{Plain Language Summary}
A Plain Language Summary (PLS) can be an incredibly effective science
communication tool. By summarizing your paper in non-technical terms,
you can explain your research and its relevance to a much broader
audience. A PLS is required for submissions to AGU Advances, G-Cubed,
GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets,
JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of
Geophysics, but optional for other journals. A PLS should be no longer
than 200 words and should be free of jargon, acronyms, equations, and
any technical information that would be unknown to people from outside
your scientific discipline. Read our tips for creating an effective PLS.

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

River runoff is an important component of the global water cycle as it
comprises about one third of the precipitation over land areas (Hagemann
et al., 2020). For the Baltic Sea river runoff is even more important,
as the Baltic Sea is nearly decoupled from the open ocean (see Figure).
As a consequence, the salinity of the Baltic Sea is driven to a large
part by freshwater supply from rivers. More generally, the freshwater
input into the Baltic Sea comes either as river runoff or a positive net
precipitation (precipitation minus evaporation) over the sea surface.
The net precipitation accounts for 11 \% and the river input for 89 \%
of the total freshwater input (Meier and Doescher, 2002). Modeling the
Baltic Sea is therefore to a large part the result of the quality of the
river input, that is used for the simulation.

In the context of climate change studies, river runoff is usually
generated in two ways. First, river runoff as input for ocean models can
be created using hyrdological models such as the Hydrological Discharge
(HD) model (Hagemann et al., 2020). HD calculates the water balance
using hydrological processes (e.g.~snow, glaciers, soil moisture,
groundwater contribution). It represents a complex forecast tool that
uses underlying physical processes. A different approach would use
data-based models that intergrate statistical correction, using the land
surface schemes of global or regional climate models.

The releatively recent rise of machine learning (ML) models has been
mostly explored for river runoff forcasting, as accurate runoff
forecasting, especially over extended periods, is pivotal for effective
water resources management (Fang et al., 2019; Tan et al., 2018; Yang et
al., 2018). Common approach us employ artificial neural networks,
support vector machines, adaptive neuro-fuzzy inference systems, and
notably, Long Short-Term Memory (LSTM) neural networks and have gained
traction for long-term hydrological forecasting due to their commendable
performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017,
Yuan et al 2018, Xu et al 2021).

LSTM networks, an evolution of the classical Recurrent Neural Networks
(RNNs), have shown stability and efficacy in sequence-to-sequence
predictions, such as using climatic indices for rainfall estimation or
long-term hydrological forecasting. However, a limitation of LSTMs is
their inability to effectively capture two-dimensional structures, an
area where Convolutional Neural Networks (CNNs) excel.

Recognizing this, we use the ConvLSTM, which integrates the strengths of
both LSTM and CNN, to extract spatiotemporal features from atmospheric
fields for predicting river runoff in the Baltic Sea catchment,
summarized by 97 inidivual rivers.

Following the previous description of the complex hyrdological model
E-Hype - when present E-Hype should be used. However, in absence of a
fully functioning hydrogolocial model, that also uses a rather complex
parametrization, RNNs represent a robust way to predict river runoff for
any give period of time using atmospheric forcing.

Modeling the Baltic Sea is to a large part the result of the quality of
the freshwater input, that is used for the simulation. Meier and
Kauker(2003) showed that decadal salinity variations of about 1 \(g\)
\(kg^{-1}\) are caused, inter alia, by annual runoff variations.
Further, Meier and Kauker (2003) showed that about 50 \% of the decadal
salinity variability can be explained by variations in freshwater input
into the Baltic Sea.

This paper delves into the application of deep learning, particularly
ConvLSTM, to the challenging task of precipitation nowcasting, a domain
yet to fully harness the potential of advanced machine learning
techniques. We present ConvLSTM as a novel solution to this
spatiotemporal sequence forecasting challenge, highlighting its
advantages and potential future applications.

@@

@@

@@

>>>>

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting},
  pdfauthor={Florian Börgel; Sven Karsten},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Climate Dynamics}

@@

\begin{document}
\title{From Precipitation to Prediction: Integrating ConvLSTM Models for
Comprehensive River Runoff Forecasting}

@@

\begin{abstract}
a
\end{abstract}

\section*{Plain Language Summary}
b

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

River runoff is an important component of the global water cycle as it
comprises about one third of the precipitation over land areas (Hagemann
et al., 2020). In the context of climate change studies, river runoff is
usually generated in two ways. First, river runoff as input for ocean
models can be created using hyrdological models such as the Hydrological
Discharge (HD) model (Hagemann et al., 2020). HD models calculates the
water balance using hydrological processes (e.g.~snow, glaciers, soil
moisture, groundwater contribution). It represents a complex forecasting
tool that uses underlying physical processes. A different approach would
use data-based models that intergrate statistical correction, using the
land surface schemes of global or regional climate models.

The releatively recent rise of machine learning (ML) models has been
mostly explored for river runoff forecasting, as accurate runoff
forecasting, especially over extended periods, is pivotal for effective
water resources management (Fang et al., 2019; Tan et al., 2018; Yang et
al., 2018). Common approaches employ artificial neural networks, support
vector machines, adaptive neuro-fuzzy inference systems, and notably,
Long Short-Term Memory (LSTM) neural networks that have gained traction
for long-term hydrological forecasting due to their excellent
performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017,
Yuan et al 2018, Xu et al 2021).

LSTM networks, an evolution of the classical Recurrent Neural Networks
(RNNs), have shown stability and efficacy in sequence-to-sequence
predictions, such as using climatic indices for rainfall estimation or
long-term hydrological forecasting. However, a limitation of LSTMs is
their inability to effectively capture two-dimensional structures, an
area where Convolutional Neural Networks (CNNs) excel. Reconizing this
limitation (SHI et al., 2015) proposed a convolutional LSTM (ConvLSTM)
architectures, which combines the strengths of both LSTM and CNN. The
ConvLSTM network has been proven useful for precipitation nowcasting
(SHI et al., 2015) or for river runoff forecasting {[}Ha et al. (2021);
@niu2020{]}.

In the following we will show that in absence of a fully functioning
hydrogolocial model, that also uses a rather complex parametrization,
ConvLSTM also represent a robust way to predict multiple rivers at once
for any given period using only atmospheric forcing. In this work we use
the Baltic Sea catchment to illustrate our approach, while in principal
the methodology we propose is universally applicable across various
geographic regions. The Baltic Sea serves as a challenging example due
to its unique hydrological characteristics, being nearly decoupled from
the open ocean (see Figure). As a consequence, the salinity of the
Baltic Sea is driven to a large part by freshwater supply from rivers.
More generally, the freshwater input into the Baltic Sea comes either as
river runoff or a positive net precipitation (precipitation minus
evaporation) over the sea surface. The net precipitation accounts for 11
\% and the river input for 89 \% of the total freshwater input (Meier
and Doescher, 2002). Modeling the Baltic Sea is therefore to a large
part the result of the quality of the river input, that is used for the
simulation. This makes the accurate modeling of river runoff especially
critical for simulations pertaining to the Baltic Sea.

In this work we will, we present a ConvLSTM architecture that is able to
predict daily river runoff for 97 rivers across the Baltic Sea
catchment.

@@

\hypertarget{runoff-data-used-for-training}{%
\subsection{Runoff data used for
training}\label{runoff-data-used-for-training}}

\hypertarget{atmospheric-forcing}{%
\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-ha2021}{}}%
Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
streamflow based on deep learning neural network with El
Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}

@@

\leavevmode\vadjust pre{\hypertarget{ref-shi2015}{}}%
SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
Convolutional LSTM network: A machine learning approach for
precipitation nowcasting. In (Vol. 28). Curran Associates, Inc.
Retrieved from
\url{https://proceedings.neurips.cc/paper/2015/hash/07563a3fe3bbe7e3ba84431ad9d055af-Abstract.html}
